\chapter{Experiments}
\label{chap:experiments}

This chapter records the current experimental protocol and main results. Full per-scene tables, expanded ablations, and finalized plots will be added after the remaining experiments are completed.

\section{Dataset and Metrics}

The current evaluation uses the nine Mip-NeRF 360 scenes: bicycle, flowers, garden, stump, treehill, room, counter, kitchen, and bonsai. The reported metrics are training time, wall time, PSNR, SSIM, paper-style VGG LPIPS, and final Gaussian count. Unless stated otherwise, the main pipeline uses COLMAP/SfM initialization and does not include optional AnySplat initialization.

\section{Main Result}

Table~\ref{tab:main-result} summarizes the current nine-scene mean results under the local evaluation protocol.

\begin{table}[H]
\centering
\caption{Current nine-scene mean comparison.}
\label{tab:main-result}
\begin{tabular}{lrrrrr}
\toprule
Method & Train time & PSNR & SSIM & LPIPS & Gaussians \\
\midrule
Original 3DGS & 1053.54s & 27.5963 & 0.8153 & 0.2138 & 2.625M \\
Local fast-training reference & 235.93s & 27.9330 & 0.8200 & 0.2160 & 1.161M \\
RapidGS & 152.47s & 28.1586 & 0.8289 & 0.2073 & 1.086M \\
\bottomrule
\end{tabular}
\end{table}

Compared with the original 3DGS baseline, RapidGS reduces average training time by approximately $6.9\times$. Compared with the local FastGS-quality reference, it reduces training time by approximately 35.4 percent while improving PSNR, SSIM, and LPIPS in the current mean result.

These numbers should be interpreted as the current main result rather than the final experimental claim. The final thesis should include per-scene values, repeated runs where practical, and runtime breakdowns. The mean table is still useful at this stage because it fixes the quality target for the rest of the draft: the proposed system should be faster than the local fast-training reference while preserving the same reconstruction-quality region.

\section{Planned Ablations}

The final thesis will separate the contribution of each cost source. The planned ablation axes are:

\begin{itemize}
  \item structure control: score-gated densification, VCP pruning, score-view budget, and score thresholds;
  \item training schedule: 30k, 27k, 25k, 23k, 22k, and 21k iterations under the same quality protocol;
  \item rasterization work: compact-box multiplier, compact rasterization off, and screen-radius pruning;
  \item backend execution: fused CUDA backward and Adam update versus less specialized optimizer paths;
  \item initialization: COLMAP/SfM initialization versus optional AnySplat initialization with initialization time reported separately.
\end{itemize}

\section{Runtime Indicators to Report}

The final experimental chapter should report intermediate cost indicators in addition to end-to-end time. Gaussian count is necessary but not sufficient. Tile-instance count, forward preprocessing time, sort time, blend time, backward/update time, score-render time, and mutation time give a clearer picture of why a configuration is faster or slower.

The most useful plots are expected to be:
\begin{itemize}
  \item Gaussian count over training iterations;
  \item effective Gaussian-tile instance count over training iterations;
  \item per-step runtime breakdown before and after compact rasterization;
  \item quality versus training-time curves for schedule choices;
  \item score-render overhead compared with the time saved by controlled structure growth.
\end{itemize}

\section{Evaluation Boundaries}

The evaluation should keep initialization and training-loop time separate. COLMAP/SfM initialization is part of the standard data preparation path for the optimization-style baseline. Optional external Gaussian initialization has its own runtime and coordinate-alignment cost, so it should be reported as an additional pipeline component.

The evaluation should also keep metric implementations fixed. PSNR, SSIM, and VGG LPIPS are reported together because each metric emphasizes a different part of reconstruction quality. A configuration that improves one metric while substantially increasing time or degrading another metric should be treated as a trade-off, not as an unconditional improvement.

\section{Current Interpretation}

The current results suggest that the speedup is not caused by a single switch. Controlled structure editing prevents uncontrolled Gaussian growth, compact rasterization reduces low-value tile work, the fused backend lowers optimizer-path overhead, and the 23k schedule reduces the total number of training steps. The final experimental chapter should therefore report both total training time and intermediate cost indicators such as Gaussian count, tile-instance count, and runtime breakdown.
