\chapter{Introduction}
\label{chap:introduction}

Novel view synthesis aims to reconstruct a scene from posed images and render photorealistic images from unseen camera viewpoints. 3D Gaussian Splatting (3DGS) has become an effective representation for this task because it combines an explicit 3D primitive set with a differentiable rasterizer. A trained 3DGS scene is a set of anisotropic 3D Gaussians with position, covariance, opacity, and view-dependent color parameters. During rendering, the Gaussians are projected to the image plane, sorted by depth within screen-space tiles, and composited with front-to-back alpha blending.

The same properties that make 3DGS attractive for rendering also make its training pipeline sensitive to system cost. The model usually starts from sparse structure and grows through repeated densification. Gaussian count can increase rapidly, especially near occlusion boundaries, high-frequency texture, and regions with unstable gradients. Every additional Gaussian increases the work of projection, tile assignment, sorting, blending, backward propagation, optimizer update, and state maintenance. Standard training also exposes large per-Gaussian gradients to PyTorch and relies on a general optimizer path, which is robust but not specialized for a dynamic Gaussian set.

This thesis focuses on fast single-scene 3DGS training without changing the core rendering representation. The goal is not to introduce a new primitive, a new supervision signal, or a new alpha compositing rule. Instead, the goal is to make the standard 3DGS training semantics execute faster and with better control over redundant structure.

\section{Training-Cost View}

For a fixed dataset, image scale, and quality target, the total training time can be approximated as
\begin{equation}
T_{\mathrm{train}}
\approx
\bar{N}_{\mathrm{gs}}
\cdot
c_{\mathrm{opt}}
\cdot
N_{\mathrm{iter}},
\end{equation}
where $\bar{N}_{\mathrm{gs}}$ is the average number of Gaussians during training, $c_{\mathrm{opt}}$ is the average per-Gaussian cost of one training step, and $N_{\mathrm{iter}}$ is the number of training iterations. This view is only a first-order approximation, because rasterization cost also depends on screen coverage, tile-instance count, depth distribution, and memory behavior. Nevertheless, it gives a useful design principle: fast training should reduce redundant structure, lower the cost of each active primitive, and avoid unnecessary optimization steps.

FasterGSFusedRapid follows this principle. First, it controls Gaussian growth and deletion through a multi-view structure-editing view. Densification and pruning are treated as discrete edits to the Gaussian set. Growth should be applied where additional capacity is likely to reduce multi-view reconstruction error, and deletion should remove Gaussians that have low contribution, abnormal scale, or persistent inconsistency. Second, it reduces the cost of one training step by fusing Gaussian backward propagation and Adam update into a CUDA backend. PyTorch still computes the RGB loss, but the bridge between the loss and the Gaussian optimizer is the image-space gradient. Third, it reduces rasterization work through compact tile-instance generation, bucketed backward execution, Morton ordering, and explicit state synchronization.

\section{Scope}

The main pipeline in this thesis uses COLMAP/SfM initialization and RGB-only supervision. Optional AnySplat Gaussian PLY initialization is treated as a separate initialization path and later ablation factor. This distinction matters because a feed-forward initializer has its own runtime, coordinate-alignment requirements, and failure modes. The main result should therefore be read as an acceleration of the standard optimization-style 3DGS training pipeline, not as a system that depends on an external prior.

This version of the thesis draft intentionally leaves several parts for later completion. Related work is represented as a chapter scaffold. Full per-scene tables, ablation results, formal figures, and expanded citation discussion will be added after the remaining experiments are finalized.

\section{Contributions}

This thesis makes three main contributions.

\begin{enumerate}
  \item It introduces a cost-aware structure-control perspective for 3DGS training. Multi-view high-error contribution statistics gate densification and support later pruning, reducing redundant growth driven by single-view gradient spikes.
  \item It presents a fused CUDA training backend that preserves the PyTorch-side RGB loss while moving Gaussian backward propagation, optimizer update, densification statistics, and dynamic state maintenance into a specialized per-Gaussian execution path.
  \item It integrates compact rasterization and a short quality-constrained schedule to reduce tile-instance work and total iteration count while preserving standard 3DGS representation and rendering semantics.
\end{enumerate}
