\chapter{Related Work}
\label{chap:related-work}

RapidGS is positioned at the intersection of neural rendering, explicit radiance-field representations, and fast 3DGS training systems. This chapter reviews the work most relevant to that position. The goal is not to survey every 3DGS variant, but to clarify which ideas motivate this thesis and which choices distinguish RapidGS from prior methods.

\section{Neural Radiance Fields and Novel View Synthesis}

Novel view synthesis reconstructs a scene from a set of images and renders it from unseen viewpoints. Neural Radiance Fields (NeRF) established a widely used formulation in which a scene is represented as a continuous volumetric function parameterized by a neural network \cite{MildenhallECCV2020NeRF}. Given a 3D position and viewing direction, the network predicts density and view-dependent radiance, and the final pixel color is obtained by differentiable volume rendering along camera rays. This formulation produced high-quality view synthesis and provided a clean optimization objective, but the repeated neural-network queries and volumetric samples made both training and rendering expensive.

Subsequent work improved the efficiency and quality of neural radiance fields by changing sampling, parameterization, or scene representation. Mip-NeRF 360 extended the NeRF family to unbounded real-world scenes and introduced anti-aliased rendering ideas that are important for the Mip-NeRF 360 benchmark used in this thesis \cite{BarronCVPR2022MipNeRF360}. Instant neural graphics primitives accelerated neural field optimization with a multiresolution hash encoding, showing that a carefully designed spatial data structure can reduce the cost of fitting neural graphics primitives \cite{MullerTOG2022InstantNGP}. Plenoxels removed the neural network from the radiance-field representation and optimized a sparse voxel grid directly, further demonstrating the value of explicit, GPU-friendly scene parameters \cite{FridovichKeilCVPR2022Plenoxels}.

These works provide two important lessons for RapidGS. First, high-quality view synthesis depends on a representation that can model geometry, visibility, and view-dependent appearance. Second, the representation's data layout and execution path strongly affect training speed. RapidGS follows this second lesson closely: it does not change the view-synthesis objective, but it redesigns how a standard 3DGS model is optimized and updated.

\section{3D Gaussian Splatting}

3D Gaussian Splatting (3DGS) replaced expensive ray marching with explicit anisotropic Gaussian primitives \cite{KerblTOG2023GaussianSplatting}. It renders those primitives with a differentiable tile-based rasterizer. A 3DGS scene stores Gaussian means, covariance parameters, opacity, and SH color coefficients. Rendering projects the Gaussians into screen space, assigns them to tiles, sorts them by depth, and alpha-composites their contributions. This explicit representation gives real-time rendering after training and substantially reduces optimization time compared with earlier neural radiance-field pipelines.

The original 3DGS training pipeline also introduced adaptive density control. Starting from sparse SfM points, the model repeatedly clones, splits, and prunes Gaussians while optimizing continuous parameters. This dynamic structure is essential for quality: sparse initialization alone cannot represent the geometry and appearance of a complex scene. However, the same mechanism creates a major systems challenge. The number of Gaussians changes over training, and each structural mutation must keep parameter tensors, optimizer state, and rasterization buffers consistent.

RapidGS keeps the core 3DGS formulation unchanged. It uses the same kind of Gaussian primitives, RGB photometric supervision, SH color, opacity, projection, tile sorting, and alpha blending semantics. The contribution is therefore not a new representation. Instead, RapidGS targets the training system around standard 3DGS: structure edits should be more selective, tile-instance generation should avoid low-contribution work, and the backward/optimizer path should be specialized for the dynamic Gaussian set.

\section{Resource-Aware Gaussian Control}

A central theme in recent 3DGS research is that many Gaussians are redundant or poorly allocated. Taming 3DGS studies 3DGS under resource constraints and uses budgeted, score-guided densification to control model size while improving training and rendering efficiency \cite{MallickSIGGRAPHAsia2024Taming3DGS}. Mini-Splatting approaches the problem from the spatial distribution of Gaussians, arguing that inefficient clustering limits quality and speed, and proposes densification and simplification strategies to obtain a more compact representation \cite{Fang2024MiniSplatting}. Mini-Splatting2 further emphasizes early aggressive densification followed by simplification and optimization, showing that the timing of capacity allocation can affect convergence speed \cite{Fang2024MiniSplatting2}.

These methods motivate RapidGS's view that density control is not only a quality mechanism but also a runtime mechanism. A Gaussian added early in training can increase forward rendering, backward propagation, optimizer update, and memory cost for many later iterations. A Gaussian pruned too late may have already consumed substantial training time. Therefore, the average Gaussian count during optimization matters, not only the final model size.

RapidGS differs from pure model-size reduction or fixed-budget approaches in its primary objective. It does not aim to produce the smallest possible final representation. Instead, it aims to reduce training time while preserving the standard reconstruction-quality target. The multi-view high-error contribution score is used as a training-time structure signal: growth should be assigned to Gaussians that participate in stable multi-view residual regions, while pruning should remove low-value or inconsistent capacity after the representation has had time to stabilize.

\section{Fast 3DGS Training Systems}

Several recent systems directly target 3DGS training speed. DashGaussian frames the optimization cost in terms of rendering resolution and primitive count, and schedules these quantities so the model gradually fits higher-frequency image content while limiting unnecessary early computation \cite{ChenCVPR2025DashGaussian}. This connects closely to the cost decomposition used in this thesis: training time depends on both the number of active primitives and the cost of each rendered step.

FastGS focuses on controlling Gaussian redundancy through multi-view consistency and redesigns densification and pruning to reduce unnecessary Gaussian growth while maintaining quality \cite{Ren2025FastGS}. Its analysis is particularly relevant because it identifies adaptive density control as a major bottleneck in vanilla 3DGS. RapidGS shares the broad motivation that single-view or weakly constrained growth can create redundant Gaussians. The difference is that RapidGS combines this structure-control direction with a fused CUDA training backend and compact rasterization path, so the method acts on both Gaussian count and per-step execution cost.

Faster-GS analyzes backend and algorithmic optimization opportunities in 3DGS training \cite{Hahlbohm2026FasterGS}. It emphasizes that fair acceleration comparisons should separate fidelity-preserving backend work from algorithmic changes that affect quality. This perspective matches the scope of RapidGS. RapidGS explicitly keeps the representation and RGB objective fixed, and treats backend fusion, compact tile assignment, and dynamic state maintenance as part of the training system rather than as changes to the scene representation.

\section{Rasterization Work and Primitive Pruning}

The cost of 3DGS training and rendering is not determined only by the number of Gaussians. It also depends on how many Gaussian-tile instances are generated and processed. Speedy-Splat targets rendering speed by reducing both sparse pixels and sparse primitives, including tighter localization of Gaussian influence and pruning strategies integrated into the pipeline \cite{Hanson2024SpeedySplat}. Its analysis supports a key observation used in RapidGS: conservative tile assignment can create unnecessary work when a projected Gaussian's bounding region includes tiles where the actual contribution is negligible.

RapidGS's compact rasterization follows the same broad principle of avoiding low-value rasterization work, but it is used inside the training pipeline together with fused backward and optimizer update. The contribution-aware compact test is designed not to change alpha blending semantics. It only decides whether a Gaussian-tile pair should enter the expensive sort, blend, and backward path. This is complementary to pruning: pruning reduces the primitive set, while compact rasterization reduces the tile work generated by primitives that remain active.

\section{Fused Backends and Dynamic State}

3DGS training differs from standard neural-network training because the parameter set changes during optimization. Clone, split, prune, and reorder operations must update Gaussian parameters, Adam moments, gradient accumulators, pruning counters, and auxiliary buffers consistently. A general PyTorch optimizer path is flexible but treats parameter tensors and optimizer state through a generic interface. For millions of per-Gaussian parameter rows, repeated gradient materialization, optimizer dispatch, and state maintenance can become a significant fixed cost.

Prior fast-training systems show that implementation details matter for 3DGS efficiency \cite{MallickSIGGRAPHAsia2024Taming3DGS,Hahlbohm2026FasterGS}. RapidGS continues in this direction by using a narrow Python/CUDA boundary: PyTorch computes the RGB loss, but CUDA consumes the image-space gradient and performs Gaussian backward propagation and Adam update in the backend. This design is not a change in optimization objective. It is a specialization of the execution path for the aligned per-Gaussian data structure.

\section{Initialization Priors}

Most optimization-style 3DGS pipelines initialize from COLMAP/SfM points and then optimize one scene at a time. Feed-forward Gaussian methods pursue a different direction: predict Gaussian primitives, camera quantities, or scene structure from input images without running a full per-scene optimizer. AnySplat is an example of this line of work, predicting 3D Gaussian primitives from unconstrained multi-view inputs in a feed-forward manner \cite{Jiang2025AnySplat}.

RapidGS treats such priors as optional initialization components rather than as its main contribution. If a feed-forward initializer provides a better Gaussian set, the subsequent 3DGS training loop can still refine it using the same rasterizer and optimizer. However, the initializer has its own runtime, coordinate-frame assumptions, and failure modes. For fair evaluation, initialization time and training-loop time should be reported separately. This is why the main RapidGS pipeline uses COLMAP/SfM initialization and frames AnySplat-style priors as a separate experimental axis.

\section{Summary}

The related work suggests three design requirements for fast 3DGS training. First, the representation should preserve the quality advantages of standard 3DGS. Second, density control should regulate redundant Gaussian growth during training, not only compress the final model. Third, the execution backend should reduce per-step overhead from rasterization, backward propagation, optimizer update, and dynamic state maintenance. RapidGS follows these requirements by combining multi-view structure control, compact tile-instance generation, fused CUDA backward/update, and a quality-constrained short schedule.
