\begin{abstract}

3D Gaussian Splatting (3DGS) represents a scene with explicit anisotropic Gaussian primitives and renders novel views through differentiable rasterization. Although this representation delivers high visual quality, standard 3DGS training remains expensive because it repeatedly grows and prunes a large Gaussian set, rasterizes many low-contribution Gaussian-tile pairs, and updates millions of Gaussian parameters through a general-purpose optimizer path.

This thesis studies how to accelerate single-scene 3DGS training while preserving the Gaussian representation, spherical-harmonic color model, front-to-back alpha-blending rule for retained instances, and RGB photometric objective. The central observation is that total time is determined by the number of training iterations and the evolving cost of each step, which depends jointly on Gaussian count, Gaussian--tile instances, rasterization, backward propagation, and optimizer execution. Based on this conceptual decomposition, the proposed system, RapidGS, reduces redundant Gaussian growth through multi-view structure control, lowers tile and backend work through compact rasterization and fused CUDA execution, and uses a shorter schedule under an explicit quality constraint.

The main RapidGS pipeline uses COLMAP/SfM initialization and an RGB-only training objective. Optional Gaussian initialization from AnySplat is treated as a separate initialization path rather than a required component of the main result. Under the evaluation protocol used in this thesis, RapidGS reduces average training time on the nine-scene Mip-NeRF 360 benchmark from 1053.54 seconds for the original 3DGS baseline to 152.47 seconds, while improving PSNR, SSIM, and LPIPS. Compared with FastGS, it reduces training time by approximately 35.4 percent while maintaining or improving reconstruction quality.

\end{abstract}
